Introduction to tableone
==================

```{r, message = FALSE, tidy = FALSE, echo = F}
## knitr configuration: http://yihui.name/knitr/options#chunk_options
library(knitr)
showMessage <- FALSE
showWarning <- TRUE
set_alias(w = "fig.width", h = "fig.height", res = "results")
opts_chunk$set(comment = "", error= TRUE, warning = showWarning, message = showMessage,
               tidy = FALSE, cache = F, echo = T,
               fig.width = 7, fig.height = 7, dev.args = list(family = "sans"))

## R configuration
options(width = 116, scipen = 5)
```

What is tableone?
--------------------
The tableone package is an R package that eases the construction of "Table 1", *i.e.*, patient baseline characteristics table commonly found in biomedical research papers. The packages can summarize both continuous and categorical variables mixed within one table. Categorical variables can be summarized as counts and/or percentages. Continuous variables can be summarized in the "normal" way (means and standard deviations) or "nonnormal" way (medians and interquartile ranges).

Load packages
--------------------
```{r}
## tableone package itself
library(tableone)
## survival pcakge for Mayo Clinic's PBC data
library(survival)
data(pbc)
```

Single group summary
--------------------


**Simplest use case**

The simplest use case is summarizing the whole dataset. You can just feed in the data frame to the main workhorse function CreateTableOne(). You can see there are 418 patients in the dataset.
```{r}
CreateTableOne(data = pbc)
```

**Categorical variable conversion**

Most of the categorical variables are coded numerically, so we either have to transform them to factors in the dataset or use factorVars argument to transform them on-the-fly. Also it's a better practice to specify which variables to summarize by the vars argument, and exclude the ID variable(s). How do we know which ones are numerically-coded categorical variables? Please check your data dictionary (in this case help(pbc)). This time I am saving the result object in a variable.

```{r}
## Get variables names
dput(names(pbc))
## Vector of variables to summarize
myVars <- c("time", "status", "trt", "age", "sex", "ascites", "hepato",
          "spiders", "edema", "bili", "chol", "albumin", "copper", "alk.phos",
          "ast", "trig", "platelet", "protime", "stage")
## Vector of categorical variables that need transformation
catVars <- c("status", "trt", "ascites", "hepato",
             "spiders", "edema", "stage")
## Create a TableOne object
tab2 <- CreateTableOne(vars = myVars, data = pbc, factorVars = catVars)
```

OK. It's more interpretable now. Binary categorical variables are summarized as counts and percentages of the second level. For example, if it is coded as 0 and 1, the "1" level is summarized. For 3+ category variable all levels are summarized. Please bear in mind, the percentages are calculated after excluding missing values.

```{r}
tab2
```


**Showing all levels for categorical variables**

If you want to show all levels, you can use showAllLevels argument to the print() method.

```{r}
print(tab2, showAllLevels = TRUE)
```

**Detailed information including missingness**

If you need more detailed information including the number/proportion missing. Use the summary() method on the result object. The continuous variables are shown first, and the categorical variables are shown second.

```{r}
summary(tab2)
```

**Summarizing nonnormal variables**

It looks like most of the continuous variables are highly skewed except time, age, albumin, and platelet (biomarkers are usually distributed with strong positive skews). Summarizing them as such may please your future peer reviewer(s). Let's do it with the nonnormal argument to the print() method. Can you see the difference. If you just say nonnormal = TRUE, all variables are summarized the "nonnormal" way.

```{r}
biomarkers <- c("bili","chol","copper","alk.phos","ast","trig","protime")
print(tab2, nonnormal = biomarkers)
```

**Fine tuning**

If you want to fine tune the table further, please check out ?print.TableOne for the full list of options.


Multiple group summary
--------------------

Often you want to group patients and summarize group by group. It's also pretty simple. Grouping by exposure categories is probably the most common way, so let's do it by the treatment variable. According to ?pbc, it is coded as (1) D-penicillmain (it's really "D-penicillamine"), (2) placebo, and (NA) not randomized. NA's do not function as a grouping variable, so it is dropped.

```{r}
tab3 <- CreateTableOne(vars = myVars, strata = "trt" , data = pbc, factorVars = catVars)
tab3
```

**Testing**

When there are two or more groups group comparison p-values are printed along with 


--------------------
- github repo: https://github.com/kaz-yos/tableone
